Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2012-02-07 boettiger.tex
spelling, formatting, linenumbers working with align environments

% erg, a bit mor transition needed?

We discuss the limitations of current approaches that rely on summary statistics
and introduce a model based approach to detect early warning signals.
We illustrate how stochastic differential equation (SDE) models can be used 
to reflect the uncertainty inherent in the detection of early warning signals.
We caution against paradigms that are not useful for capturing this uncertainty in 
a model-selection based approach, such as information criteria,
and suggest reciever-operating characteristics~\citep{Green1989, Keller2009}
as a powerful way to illustrate the sensitivity of different data sets 
and different early warning indicators have in detecting early warning signals.

@@

A basic difficulty arises from the need to assume a time-series is ergodic 
while trying to test if it is not.  
Theoretically, the increasing trend in variance, autocorrelation, or other statistics
is something that would be measured across an ensemble -- across replicates. %% citation? 
As true replicates are seldom avialable in systems for which developing warning signals would be most desirable,
these methods average across a single replicate using a moving window in time.  
The selection of the size of this window and whether and by how much to overlap consecutive windows
varies across the literature.  
\citet{Lenton2012} demonstrates that these differences can influence the results, 
and that the different choices each carry advantages and disadvantages.

@@

\subsubsection*{Problematic null models}
Specifying an appropriate null model is also difficult.  
Non-parametric null hypotheses seem to require the fewest assumptions but in fact can be the most problematic.  
For instance, the standard non-parametric hypothesis test with Kendall's tau rank correlation coefficient
assumes only that the two variables are independent,
but this is an assumption that is violated by the very experimental design:
temporal correlations will exist in any finely-enough samled time series, 
and moving windows introduce temporal correlations in the statistics.
Under such a test any adequately large data set will find a significant result,
regardless of whether a warning signal exists.
A similar problem arises when the points in the time series are reordered to create a null hypothesis -- 
this destroys the natural autocorrelation in the time series.  
More promising parametric null models have been proposed, 
such as autoregressive models in~\citet{Dakos2008}, bringing us  
closer to a model-based approach with explicit assumptions.  
Others have looked for alternative summary statistics where 
reasonable null models are more readily available, 
such as~\citet{Seekell2011}'s proposal to test for conditional heteroscedaticity.

@@

In this section we outline how the detection of early warning signals may be thought of
as a problem of model choice.
We next show generic models can be constructed under the assumptions discussed above
and estimated from the data in a maximum likelihood framework.
We highlight the disadvantages of comparing these estimates by information criterea, 
and insead introduce a simulation or bootstrapping approach rooted in~\citet{Cox1961} and~\citet{McLachlan1987} that
characterizes the rate of missed detections and false alarms expected in the estimate.

\subsection*{Early warning signals as model choice}
It may be useful to think of the detection of early warning signals 
as a problem of model choice rather than one of pattern recognition.  
The model choice approach attempts to frame each of the possible scenarios as structurally different equations,
each with unknown parameters that must be estimated from the data.
In any model choice problem, it is important to identify the goal of the exercise --
such as the ability to generalize, to immitate reality, or to predict~\citep{Levins1966}.  
In this case generality is more important than realism or predictive capability: 
we will write down a general model that is capable of approximating
a wide class of models in which regime shifts are characterized by a saddle node bifurcation,
and a second generic model that is capable of representing the behavior of such systems
when they are not approaching a bifurcation.  
These may be thought of as the hypothesis and null hypothesis,
though they are in fact compound hypotheses,
as we must first estimate the model parameters from the data.  
In this approach it is not assumed that ``reality'' is included in the models being tested per se, 
but that one of the models is a better approximation of the true dynamics than the other.  
System whose dynamics violate the assumptions common to both models, 
such as in the examples of ~\citet{Hastings2010} where systems exhibit sudden transitions without warning, 
fall outside the set of cases where this approach would be valid; 
though the inability of either model to match the system dynamics could be an indication of such a violation.

\subsection*{Models}
The normal form~\citep{Guckenheimer1983, Kuehn2011} for the saddle-node bifurcation is
\begin{equation}
\frac{\ud x}{\ud t} = r_t- x^2.
\label{saddle-node}
\end{equation}
where $x$ is the state variable and $r_t$ our bifurcation parameter, 
which may be slowly varying in time and 
consequently moving the system closer to a critical transition or regime shift~\citep{Scheffer2009}.  
Transforming this canonical form to allow for an arbitrary mean $\theta$,
the bifurcation looks like $ dx/dt = r_t- (\theta-x)^2 $, with fixed point $\hat x = \sqrt{r_t} +\theta =: \phi(r_t)$.
We expand around the fixed point and express in terms of the stochastic dynamics:

@@

As we discuss above, in this paradigme we must include an assumption on how the bifurcation parameter, $r_t$, is changing.
We assume a gradual, monotonic change which we approximate to first order: 
\begin{equation}
r_t = r_0 - m t.
\label{R_t}
\end{equation}
Detecting accelerating or otherwise nonlinear approaches to the bifurcation will generally require more power.
When the system is stable, $r_t$ is constant and Equation~\eqref{LSN} will reduce to a simple Ornstein-Uhlenbeck process, 
\begin{equation}
\ud X_t = r (\theta - X_t) \ud t + \sigma \ud B_t \label{OU}
\end{equation}
This is the continuous time analog of the first-order autoregressive model considered as a null model elsewhere~\citep[\emph{e.g.}][]{Dakos2008, Guttal2008a}.

\subsection*{Likelihood calculations}
The probability $P(X|M)$ of the data $X$ given the model $M$ is the product of the probability of observing each point in the time series given the previous point and the length of the interval,  
\begin{equation}
\log P(X | M)=  \sum_i \log P(x_i | x_{i-1}, t_i)
\end{equation}
For~\eqref{LSN} or~\eqref{OU} it is sufficient~\citep{Gardiner2009} to solve the moment equations for mean and variance respectively:
\begin{align}
 \frac{\ud }{\ud t} E(x| M)&=  f(x) \\
\frac{\ud}{\ud t} V(x| M) &=  -\partial_x f(x) V(x|M) + g(x)^2 
  \label{general_moments}
\end{align}

For the OU process, we can solve this in closed form over an interval of time $t_i$ between subsequent observations: 
\begin{align}
  E(x_i| M = \text{OU}) &= X_{i-1} e^{-r t_i} + \theta \left(1 - e^{-rt_i} \right) \\
V(x_i| M = \text{OU}) &= \frac{\sigma^2}{2 r} \left(1 - e^{-2 r t_i} \right)
\label{OUsoln}
\end{align}

For the time dependent model, we have analytic forms only for the dynamical equations of these moments from equation~\eqref{general_moments}, which we must integrate numerically over each time interval.
The moments of Equation~\eqref{LSN} are given by
\begin{align}
\frac{\ud }{\ud t} E(x_i| M = \text{LSN})&=  2\sqrt{r(t)}(\sqrt{r(t)}+\theta - x_i) \\
\frac{\ud}{\ud t} V(x_i| M = \text{LSN}) &=  -2 \sqrt{r(t)} V(x_i) + \sigma^2 ( \sqrt{r(t)}+\theta )
\label{LSNsoln}
\end{align}

@@

\subsection*{Information criteria will not serve.}
One will commonly observe models representing alternative processes being compared through 
the use of various information criteria such as the Akaike information criterion.  
While tempting to apply in this situation, such approaches are not suited to this problem for several reasons.  
The first is that information criteria are not concerned with the model choice objective we have in mind,
as they are typically applied to find an adequate model description without too many parameters that the system may be overfit.  
More pointedly, information criteria have no inherent notion of uncertainty.
Information criteria tests alone will not tell us our chances of a false alarm, of missing a real signal, 
or how much data we need to be confident in our ability to detect transitions.

\subsection*{Beyond hypothesis testing}
It is possible to frame the question of sensitivity, reliability, and adequate data in the language of hypothesis testing. 
This introduces the a nuisance parameter, known as the statistical significance criterion.  
In the hypothesis testing framework, a false positive is a Type I error, 
which is defined relative to this arbitrary statistical significance criterion, most commonly 0.05.  
By changing the criterion, one can increase or decrease the probability of the Type I error 
at the cost of decreasing or increasing false negative or Type II error, which must also be defined relative to this criterion.

The language of hypothesis testing is built around a bias that false positives are worse than false negatives, 
and consequently an a emphasis on $p$-values rather than power.  
In the context of early warning signals this is perilous -- 
it suggests that we would rather fail to predict a catastrophe than to sound a false alarm.

To avoid this linguistic bias and the introduction of an nuisance parameter on which to define statistical significance, 
we propose the use of Reciever Operating characteristic (ROC) curves.

\subsection*{ROC Curves}
We illustrate the trade-off between false alarms and failed detection using 
receiver-operating characteristic (ROC) curves first developed in signal-processing literature~\citep{Green1989, Keller2009}⁠. 
The curves represent the corresponding false alarm rate at any detection sensitivity (true positive rate), Fig~\ref{fig:roc_intro}.
The closer these distributions are to one-another, the more severe the trade-off.  
If the distributions overlap exactly, the ROC curve has a constant slope of unity.  
The ROC curve demonstrates this trade-off between accuracy and sensitivity.  
Different early-warning indicators will vary in their sensitivity to detect differences
between stable systems and those approaching a critical transition,
making the ROC curves a natural way to compare their performance.  
Since the shape of the curve will also depend on the duration and frequency of the time-series observations,
we can use these curves to illustrate by how much a given increase in sampling effort can decrease the rate of false alarms or failed detections.  
%The area under the ROC curve provides a summary of how well the method compares to a random guess --
%an area of 0.5 corresponding to perfectly overlapping distributions is performing no better than chance,
%while a statistic with area of unity always distinguishes correctly between stable systems and those approaching a transition.

@@

This corresponds to the well-studied ecosystem model of over-exploitation~\citep{Noy-Meir1975, May1977}, 
with stochasticity introduced from directly through the demographic process.   
This model is forced through a bifurcation by gradually increasing the $a$ parameter,
which increases can be thought of as an increasing toxicity of the environmnent
(from $a_0 = 100$ increasing at constant rate of 0.09 units/unit time).  
Other parameters are: $Xo = 730$, $e = 0.5$, $K = 1000$, $h = 200$.  
We run this model over a time interval from 0 to 500.  
and sample at 40 evenly spaced time points, which were used for subsequent analysis.
We refer to this as the simulation under a degrading environment.
We repeat this simulation with constant value of $a$, and which we refer to as the simulation in a stable environment.  
For the convenience of the inquisitive reader, we have also provided a simple function in the 
associated R package where the user can vary the sampling scheme and parameter values and rerun this analysis.

@@

Our second empirical data set comes from the glaciation record seen in deuterium levels in Antartic ice cores~\citep{Petit1999}, 
as analyzed by~\citet{Dakos2008}.
The data is preprocessed by linear interpolation and de-trending by Gaussian kernel smoothing 
to be as consistent as possible with the original analysis.
We focus on the third glaciation event, consisting of 121 sample points. 
The match is not exact since estimates the de-trending window size manually,
but the estimated correlations in the first-order auto-regression coefficients are in close agreement with that analysis. 
De-trending is intended to make the data consistent with the assumptions of the warning signal detection~\citep{Dakos2008}, 
which did not apply to the other data sets~\citep{Drake2010}.

\subsection*{Analysis}
We perform the identical analysis proceedure described above on each of these three data sets.
First, we estimate parameters for the null and test model to each data set by maximum likelihood.
Comparing the likelihood of these fits directly gives us only a minimal indication of which model fits better --
the deviances $\delta$, Eqn~\eqref{delta} -- are 5.1, 6.0, 83.9 respectively for the simulation,
the chemostat data and the glaciation data. 
The size of these differences reflects not only the magnitude of the difference in fit
between the models but also the arbitrary units of the raw likelihoods,
which are smaller for larger data-sets.  
Consequently the glaciation score reflects as much the greater length of its time series as it does anything else.

@@

The model parameters of both models are re-estimated on both families of replicates 
(the null and test, \emph{i.e.} $2 \times 2 \times 500$ fits). 
The differences in the likelihood values between the model estimates produced from the first set of simulations 
determines the null distribution for the deviance statistic $\delta$.  
As the constant OU process model is nested within the time-heterogenous model, these values are always positive,
but tend to be not as large as those produced when the models are fit to the second family of data. 
These two distributions are shown in the lower panel of 
Figures~\ref{fig:simulation},~\ref{fig:chemostat} and~\ref{fig:glaciation}, 
along with the deviance observed in the original data (vertical line).

The extent to which these distributions overlap indicates our inability to distinguish between these scenarios.  
The tendancy of the observed deviance (vertical line) to fall more clearly in the domain of one distribution or the other 
indicates the probability our observed data corresponds best with that model  -- 
either approaching a critical transition or remaining stable.  
While it trivial to assign a statistical significance to this observation based 
on how far into the tail of the (red) null distribution it falls, 
for the reasons we discussed we prefer the more symmetric comparison of the probability that 
this value was observed in either distribution.  
We visualize the trade-off between false alarms and failed detection using the ROC curves introduced above.

>>>>

We discuss the limitations of current approaches that rely on summary statistics
and introduce a model based approach to detect early warning signals.
We illustrate how stochastic differential equation (SDE) models can be used 
to reflect the uncertainty inherent in the detection of early warning signals.
We caution against paradigms that are not useful for capturing this uncertainty in 
a model-selection based approach, such as information criteria,
and suggest receiver-operating characteristics~\citep{Green1989, Keller2009}
as a powerful way to illustrate the sensitivity of different data sets 
and different early warning indicators have in detecting early warning signals.

@@

A basic difficulty arises from the need to assume a time-series is ergodic 
while trying to test if it is not.  
Theoretically, the increasing trend in variance, autocorrelation, or other statistics
is something that would be measured across an ensemble -- across replicates. %% citation? 
As true replicates are seldom available in systems for which developing warning signals would be most desirable,
these methods average across a single replicate using a moving window in time.  
The selection of the size of this window and whether and by how much to overlap consecutive windows
varies across the literature.  
\citet{Lenton2012} demonstrates that these differences can influence the results, 
and that the different choices each carry advantages and disadvantages.

@@

\subsubsection*{Problematic null models}
Specifying an appropriate null model is also difficult.  
Non-parametric null hypotheses seem to require the fewest assumptions but in fact can be the most problematic.  
For instance, the standard non-parametric hypothesis test with Kendall's tau rank correlation coefficient
assumes only that the two variables are independent,
but this is an assumption that is violated by the very experimental design:
temporal correlations will exist in any finely-enough sampled time series, 
and moving windows introduce temporal correlations in the statistics.
Under such a test any adequately large data set will find a significant result,
regardless of whether a warning signal exists.
A similar problem arises when the points in the time series are reordered to create a null hypothesis -- 
this destroys the natural autocorrelation in the time series.  
More promising parametric null models have been proposed, 
such as autoregressive models in~\citet{Dakos2008}, bringing us  
closer to a model-based approach with explicit assumptions.  
Others have looked for alternative summary statistics where 
reasonable null models are more readily available, 
such as~\citet{Seekell2011}'s proposal to test for conditional heteroscedaticity.

@@

In this section we outline how the detection of early warning signals may be thought of
as a problem of model choice.
We next show generic models can be constructed under the assumptions discussed above
and estimated from the data in a maximum likelihood framework.
We highlight the disadvantages of comparing these estimates by information criteria, 
and instead introduce a simulation or bootstrapping approach rooted in~\citet{Cox1961} and~\citet{McLachlan1987} that
characterizes the rate of missed detections and false alarms expected in the estimate.

\subsection*{Early warning signals as model choice}
It may be useful to think of the detection of early warning signals 
as a problem of model choice rather than one of pattern recognition.  
The model choice approach attempts to frame each of the possible scenarios as structurally different equations,
each with unknown parameters that must be estimated from the data.
In any model choice problem, it is important to identify the goal of the exercise --
such as the ability to generalize, to imitate reality, or to predict~\citep{Levins1966}.  
In this case generality is more important than realism or predictive capability: 
we will write down a general model that is capable of approximating
a wide class of models in which regime shifts are characterized by a saddle node bifurcation,
and a second generic model that is capable of representing the behavior of such systems
when they are not approaching a bifurcation.  
These may be thought of as the hypothesis and null hypothesis,
though they are in fact compound hypotheses,
as we must first estimate the model parameters from the data.  
In this approach it is not assumed that ``reality'' is included in the models being tested, 
but that one of the models is a better approximation of the true dynamics than the other.  
System whose dynamics violate the assumptions common to both models, 
such as in the examples of ~\citet{Hastings2010} where systems exhibit sudden transitions without warning, 
fall outside the set of cases where this approach would be valid; 
though the inability of either model to match the system dynamics could be an indication of such a violation.

\subsection*{Models}
The normal form~\citep{Guckenheimer1983, Kuehn2011} for the saddle-node bifurcation is
\begin{equation}
\frac{\ud x}{\ud t} = r_t- x^2.
\label{saddle-node}
\end{equation}
where $x$ is the state variable and $r_t$ our bifurcation parameter, 
which may be slowly varying in time and 
consequently moving the system closer to a critical transition or regime shift~\citep{Scheffer2009}.  
Transforming this canonical form to allow for an arbitrary mean $\theta$,
the bifurcation looks like \( dx/dt = r_t- (\theta-x)^2 \), with fixed point \(\hat x = \sqrt{r_t} +\theta =: \phi(r_t)\).
We expand around the fixed point and express in terms of the stochastic dynamics:

@@

As we discuss above, in this paradigm we must include an assumption on how the bifurcation parameter, $r_t$, is changing.
We assume a gradual, monotonic change which we approximate to first order: 
\begin{equation}
r_t = r_0 - m t.
\label{R_t}
\end{equation}
Detecting accelerating or otherwise nonlinear approaches to the bifurcation will generally require more power.
When the system is stable, $r_t$ is constant and Equation~\eqref{LSN} will reduce to a simple Ornstein-Uhlenbeck process, 
\begin{equation}
\ud X_t = r (\theta - X_t) \ud t + \sigma \ud B_t \label{OU}
\end{equation}
This is the continuous time analog of the first-order autoregressive model considered as a null model elsewhere~\citep[\emph{e.g.}][]{Dakos2008, Guttal2008a}.

\subsection*{Likelihood calculations}
The probability \(P(X|M) \) of the data \(X\) given the model \(M\) is the product of the probability of observing each point in the time series given the previous point and the length of the interval,  
\begin{equation}
\log P(X | M)=  \sum_i \log P(x_i | x_{i-1}, t_i)
\end{equation}

For~\eqref{LSN} or~\eqref{OU} it is sufficient~\citep{Gardiner2009} to solve the moment equations for mean and variance respectively:

\begin{align}
 \frac{\ud }{\ud t} E(x| M)&=  f(x) \\
\frac{\ud}{\ud t} V(x| M) &=  -\partial_x f(x) V(x|M) + g(x)^2 
  \label{general_moments}
\end{align}

For the OU process, we can solve this in closed form over an interval of time $t_i$ between subsequent observations:

\begin{align}
  E(x_i| M = \text{OU}) &= X_{i-1} e^{-r t_i} + \theta \left(1 - e^{-rt_i} \right) \\
V(x_i| M = \text{OU}) &= \frac{\sigma^2}{2 r} \left(1 - e^{-2 r t_i} \right)
\label{OUsoln}
\end{align}

For the time dependent model, we have analytic forms only for the dynamical equations of these moments from equation~\eqref{general_moments}, which we must integrate numerically over each time interval.
The moments of Equation~\eqref{LSN} are given by

\begin{align}
\frac{\ud }{\ud t} E(x_i| M = \text{LSN})&=  2\sqrt{r(t)}(\sqrt{r(t)}+\theta - x_i) \\
\frac{\ud}{\ud t} V(x_i| M = \text{LSN}) &=  -2 \sqrt{r(t)} V(x_i) + \sigma^2 ( \sqrt{r(t)}+\theta )
\label{LSNsoln}
\end{align}

@@

\subsection*{Information criteria will not serve.}
One will commonly observe models representing alternative processes being compared through 
the use of various information criteria such as the Akaike information criterion.  
While tempting to apply in this situation, such approaches are not suited to this problem for several reasons.  
The first is that information criteria are not concerned with the model choice objective we have in mind,
as they are typically applied to find an adequate model description without too many parameters that the system may be over-fit.  
More pointedly, information criteria have no inherent notion of uncertainty.
Information criteria tests alone will not tell us our chances of a false alarm, of missing a real signal, 
or how much data we need to be confident in our ability to detect transitions.

\subsection*{Beyond hypothesis testing}
It is possible to frame the question of sensitivity, reliability, and adequate data in the language of hypothesis testing. 
This introduces a nuisance parameter known as the statistical significance criterion.  
In the hypothesis testing framework, a false positive is a Type I error, 
which is defined relative to this arbitrary statistical significance criterion, most commonly 0.05.  
By changing the criterion, one can increase or decrease the probability of the Type I error 
at the cost of decreasing or increasing false negative or Type II error, which must also be defined relative to this criterion.

The language of hypothesis testing is built around a bias that false positives are worse than false negatives, 
and consequently an emphasis on $p$-values rather than power.  
In the context of early warning signals this is perilous -- 
it suggests that we would rather fail to predict a catastrophe than to sound a false alarm.

To avoid this linguistic bias and the introduction of an nuisance parameter on which to define statistical significance, 
we propose the use of receiver operating characteristic (ROC) curves.

\subsection*{ROC Curves}
We illustrate the trade-off between false alarms and failed detection using 
receiver-operating characteristic curves first developed in signal-processing literature~\citep{Green1989, Keller2009}⁠. 
The curves represent the corresponding false alarm rate at any detection sensitivity (true positive rate), Fig~\ref{fig:roc_intro}.
The closer these distributions are to one-another, the more severe the trade-off.  
If the distributions overlap exactly, the ROC curve has a constant slope of unity.  
The ROC curve demonstrates this trade-off between accuracy and sensitivity.  
Different early-warning indicators will vary in their sensitivity to detect differences
between stable systems and those approaching a critical transition,
making the ROC curves a natural way to compare their performance.  
Since the shape of the curve will also depend on the duration and frequency of the time-series observations,
we can use these curves to illustrate by how much a given increase in sampling effort can decrease the rate of false alarms or failed detections.  
%The area under the ROC curve provides a summary of how well the method compares to a random guess --
%an area of 0.5 corresponding to perfectly overlapping distributions is performing no better than chance,
%while a statistic with area of unity always distinguishes correctly between stable systems and those approaching a transition.

@@

This corresponds to the well-studied ecosystem model of over-exploitation~\citep{Noy-Meir1975, May1977}, 
with stochasticity introduced from directly through the demographic process.   
This model is forced through a bifurcation by gradually increasing the $a$ parameter,
which increases can be thought of as an increasing toxicity of the environment
(from $a_0 = 100$ increasing at constant rate of 0.09 units/unit time).  
Other parameters are: $Xo = 730$, $e = 0.5$, $K = 1000$, $h = 200$.  
We run this model over a time interval from 0 to 500.  
and sample at 40 evenly spaced time points, which were used for subsequent analysis.
We refer to this as the simulation under a degrading environment.
We repeat this simulation with constant value of $a$, and which we refer to as the simulation in a stable environment.  
For the convenience of the inquisitive reader, we have also provided a simple function in the 
associated R package where the user can vary the sampling scheme and parameter values and rerun this analysis.

@@

Our second empirical data set comes from the glaciation record seen in deuterium levels in Antarctic ice cores~\citep{Petit1999}, 
as analyzed by~\citet{Dakos2008}.
The data is preprocessed by linear interpolation and de-trending by Gaussian kernel smoothing 
to be as consistent as possible with the original analysis.
We focus on the third glaciation event, consisting of 121 sample points. 
The match is not exact since estimates the de-trending window size manually,
but the estimated correlations in the first-order auto-regression coefficients are in close agreement with that analysis. 
De-trending is intended to make the data consistent with the assumptions of the warning signal detection~\citep{Dakos2008}, 
which did not apply to the other data sets~\citep{Drake2010}.

\subsection*{Analysis}
We perform the identical analysis procedure described above on each of these three data sets.
First, we estimate parameters for the null and test model to each data set by maximum likelihood.
Comparing the likelihood of these fits directly gives us only a minimal indication of which model fits better --
the deviances $\delta$, Eqn~\eqref{delta} -- are 5.1, 6.0, 83.9 respectively for the simulation,
the chemostat data and the glaciation data. 
The size of these differences reflects not only the magnitude of the difference in fit
between the models but also the arbitrary units of the raw likelihoods,
which are smaller for larger data-sets.  
Consequently the glaciation score reflects as much the greater length of its time series as it does anything else.

@@

The model parameters of both models are re-estimated on both families of replicates 
(the null and test, \emph{i.e.} $2 \times 2 \times 500$ fits). 
The differences in the likelihood values between the model estimates produced from the first set of simulations 
determines the null distribution for the deviance statistic $\delta$.  
As the constant OU process model is nested within the time-heterogeneous model, these values are always positive,
but tend to be not as large as those produced when the models are fit to the second family of data. 
These two distributions are shown in the lower panel of 
Figures~\ref{fig:simulation},~\ref{fig:chemostat} and~\ref{fig:glaciation}, 
along with the deviance observed in the original data (vertical line).

The extent to which these distributions overlap indicates our inability to distinguish between these scenarios.  
The tendency of the observed deviance (vertical line) to fall more clearly in the domain of one distribution or the other 
indicates the probability our observed data corresponds best with that model  -- 
either approaching a critical transition or remaining stable.  
While it trivial to assign a statistical significance to this observation based 
on how far into the tail of the (red) null distribution it falls, 
for the reasons we discussed we prefer the more symmetric comparison of the probability that 
this value was observed in either distribution.  
We visualize the trade-off between false alarms and failed detection using the ROC curves introduced above.
